\documentclass{report}

% ==============================

% Configuration files for custom packages and macros
\input{config/preamble}
\input{config/macros}
\input{config/letterfonts}

% ==============================

\title{\Huge{COMP 458/558}\\Introduction to Quantum Computing}
\author{\huge{Micah Kepe}}
\date{}

\begin{document}

\maketitle
\newpage% or \cleardoublepage
\pdfbookmark[<level>]{<title>}{<dest>}
\pdfbookmark[section]{\contentsname}{toc}
\tableofcontents
\pagebreak

% ==============================
% PHASE I: INTRODUCTION
% ==============================
\chapter{Phase I: Introduction and Background}

\input{lectures/phase1/lecture1.tex}

\input{lectures/phase1/lecture2.tex}

\input{lectures/phase1/lecture3.tex}

% ==============================
% PHASE II: FUNDAMENTALS
% ==============================
\chapter{Phase II: Fundamentals of Quantum Algorithms}

\section{Lecture 4: Quantum State Representation and Qubits}
\dfn{Quantum States}{Define quantum states.}
\dfn{Qubits}{Define qubits and their representation.}

\section{Lecture 5: Quantum Gates and Transformations}
\dfn{Unitary Matrices}{Discuss unitary matrices.\footnote{A \textbf{unitary matrix} $U$ satisfies $U^{\dagger}U = I$, where $U^{\dagger}$ is the conjugate transpose of $U$.}}
\ex{Hadamard Gate}{Provide an example and matrix representation of the Hadamard gate.}

\section{Lecture 6: Quantum Circuits and Entanglement}
\dfn{Entanglement}{Define entanglement.}
\ex{Bell State}{Example of a Bell state and its significance.}

% ==============================
% PHASE III: ADVANCED
% ==============================
\chapter{Phase III: Advanced Quantum Algorithms}

\section{Lecture 7: Grover's Search Algorithm}
\dfn{Grover's Algorithm}{Overview of Grover's search algorithm.}
\qs{}{Provide a pseudocode or example implementation in Python.}

\section{Lecture 8: Variational Quantum Algorithms}
\dfn{VQE}{Introduce the Variational Quantum Eigensolver (VQE).}

% ==============================
% PHASE IV: SPECIAL TOPICS
% ==============================
\chapter{Phase IV: Special Topics in Quantum Computing}

\section{Lecture 9: Quantum Annealing}
\dfn{Quantum Annealing}{Discuss the basics of quantum annealing.}

\section{Lecture 10: Quantum Machine Learning}
\dfn{QML}{Define quantum machine learning and its applications.}

% ==============================
% PHASE V: CONCLUDING LECTURES
% ==============================
\chapter{Phase V: Concluding Lectures}

\section{Lecture 11: Project Presentations and Feedback}
\nt{Placeholder for project presentation notes.}

\end{document}
