Quantum computing is an emerging field poised to revolutionize multiple
disciplines, including cryptography, optimization, machine learning, and
scientific simulations. This handbook serves as a structured compilation of
notes from \textit{COMP 458/558: Quantum Computing Algorithms}, offered at Rice
University during Spring 2025. The content is derived from lectures,
discussions, and supplementary readings throughout the semester, aiming to
reinforce core concepts and provide a comprehensive reference for students
and enthusiasts alike.

\vspace{0.3cm}

\noindent
The notes begin with foundational mathematical principles, including linear
algebra and probability theory. Subsequent sections explore quantum circuits,
essential quantum gates, multi-qubit systems, and fundamental quantum
algorithms such as Grover’s search and Shor’s factoring algorithm. The latter
phases of the handbook introduce advanced topics, including variational
quantum algorithms, quantum simulation techniques, and emerging paradigms in
quantum machine learning and optimization.

\vspace{0.3cm}

\noindent
I would like to extend my deepest gratitude to \textbf{Professor Tirthak Patel},
whose guidance and expertise have been instrumental in shaping my
understanding of quantum computing. I am also grateful to my peers for the
collaborative discussions that enriched my learning experience.

\vspace{0.3cm}

\noindent
As quantum computing continues to evolve, these notes will remain a living
document, subject to refinement and expansion. I welcome feedback,
corrections, and contributions from readers to ensure the accuracy and
completeness of this resource.

\vspace{1cm}

\noindent \textbf{Micah Kepe}  \\
Rice University  \\
Spring 2025
